Exploratory data structure comparisons: Three new visual tools based on Principal Component Analysis

### 2018-06-23 article/Old versions/MPS/PCADSC_article.tex
Minor changes to manuscript + figure code

We assume that the variables have been standardized before PCA is conducted, which is equivalent to working with the empirical correlation matrices rather than the empirical covariance matrices. We also assume that all considered variables have a numerical interpretation, for example by being continuous or ordinally categorical. Moreover, we use the following notation: 
$\mathbf{X} \in \RR^{n_x \times d} $ and $\mathbf{Y} \in \RR^{n_y \times d} $ are datasets containing the same number of variables, $d$, but possibly different numbers of observations, $n_x$ and $n_y$. We use $X_j$ to refer to the $j$th variables of $\mathbf{X}$ and $x_{ij}$ to refer to the $i$th observation  within that variable, while $x_{i \cdot}$ is the full $i$th observation row. Note that $X_j \in \RR^{n_x}$, $x_{ij} \in \RR$, and $x_{i \cdot} \in \RR^d$. We let $\bar{X} = (\bar{X}_1, ... , \bar{X}_d)^T = \frac{1}{n_x} \sum_{i = 1} x_{i \cdot}$ denote the variable averages and use $S_x =  \frac{1}{n_x - 1} \sum_{i = i}^n (x_{i \cdot} - \bar{X}) - (x_{i \cdot} - \bar{X})^T \in \RR^{d \times d}$ to denote the empirical covariance matrix of $\mathbf{X}$.

@@

\subsection*{The angle plot}
The angle plot compares both loadings and eigenvalues at once and it can be used to understand the information loss if the data structure of one dataset is superimposed on the other, thereby revealing which principal components (i.e. loading and eigenvalue pairs) are most similar and most different across the two datasets. Let $\lambda_{\max} = \max\{ \lambda_{1}^x, \lambda_{1}^y \}$ be the largest eigenvalue for the two datasets. Then the empirical correlation matrix for the $\mathbf{X}$ dataset, $S_x$, has the following orthogonal decomposition in the coordinate system of the $\mathbf{Y}$ dataset
\begin{equation*}
S_x = \sum_{j=1}^d \lambda_{j}^x \eta_{j}^x ({\eta_{j}^x})^\top
= \lambda_{\max} \sum_{k=1}^d
\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{k}^x}{\lambda_{\max}}} \eta_{j}^y ({\eta_{j}^y}^\top \eta_{k}^x) \Bigg)
{\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{k}^x}{\lambda_{\max}}} \eta_{j}^y ({\eta_{j}^y}^\top \eta_{k}^x) \Bigg)}^\top,
\end{equation*}
and we have a similar decomposition of $S_y$ in the coordinate system of the $\mathbf{X}$ dataset. We propose to visualize these two decompositions in a $d \times d$ grid display. In the $j$th row and $k$th column of this display we plot two arrows based at the lower left corner of the grid cell. The first arrow has length $\mu_{jk}$ and angle $\theta_{jk}/2$ counterclockwise from the diagonal, and the second arrow has length $\nu_{jk}$ and angle $\theta_{jk}/2$ clockwise from the diagonal. To facilitate the following description we will refer to the arrows drawn counterclockwise as the blue arrows, and the arrows drawn clockwise as the red arrows. The lengths $\mu_{jk}$ and $\nu_{jk}$, and the angle $\theta_{jk}$, are given by
\begin{align*}
\mu_{jk} &= \sqrt{\frac{\lambda_{k}^x}{\lambda_{\max}}} \lvert {\eta_{k}^x}^\top \eta_{j}^y \rvert, &
\nu_{jk} &= \sqrt{\frac{\lambda_{j}^y}{\lambda_{\max}}} \lvert {\eta_{j}^y}^\top \eta_{k}^x \rvert, &
\theta_{jk} &= \acos(\lvert {\eta_{k}^x}^\top \eta_{j}^y \rvert).
\end{align*}
For two $d$-dimensional, unit length vectors $a$ and $b$, it holds that $ a^\top b = \langle a, b \rangle = \tilde{c}(a,b)$, where $\tilde{c}$ denotes the empirical correlation function. Thus, in the angle plot, we are essentially looking at the absolute values of correlations between loadings that have been scaled according to their variance contributions. The absolute value of the projection ${\eta_{k}^x}^\top \eta_{j}^y$ is inserted due to the indeterminacy of the direction of loading vectors. This indeterminacy implies that the angle between loadings from the two datasets always can be chosen to be in the interval $[0,\pi/2]$, and hence the decomposition of $S_x$ and $S_y$ can be visualized in a joint plot by dividing the angles by two and using counterclockwise and clockwise shifts from the diagonal. Furthermore, the scaling of the lengths by $\lambda_{\max}$ is made so that the longest arrow has at most unit length.

In the \emph{angle plot}, the blue arrows in the $k$th column of the grid display visualize the decomposition of the $k$th principal component of the $\mathbf{A}$ dataset in the coordinate system of the $\mathbf{B}$ dataset. Similarly, the red arrows in the $j$th row visualize the decomposition of the $j$th principal components of the $\mathbf{B}$ dataset in the coordinate system of the $\mathbf{A}$ dataset. \hl{If we have coinciding blue and red arrows along the diagonal in the grid display, and arrows of vanishing lengths in the off-diagonal cells, then the structures of the two datasets are identical. In contrast, differences in the eigenvalues are visualized as differences in the lengths of the blue and the red arrows, also in the diagonal. And loadings in other directions than the corresponding loading from the other dataset are visualized as angle separation of the blue and the red arrows in the diagonal cells, as well as arrows of non vanishing length in the off-diagonal cells.}

\hl{However, there is a generic situation where the angles between the arrows lack discriminative power, namely when two or more neighbouring eigenvalues are identical or close. For instance, multiple eigenvalues close to 1 arise if more than one of the variables are uncorrelated, or close to uncorrelated, of all the other variables. In this situation only the corresponding eigenspace is well estimates, but the rotation of the associated eigenvectors will be subject to sampling variability. To investigate if we are in such a situation of identical data structures with neighbouring eigenvalues being close, where comparably long arrows may occur in the corresponding blocks of off-diagonal cells, we have implemented a permutation test similar as done for the CE-plot. For each random reallocation we recompute the angles $\theta_{jk}$. The resampled angles may both be used for quantitative tests, where large angles are critical for the diagonal cells and small angles are critical for the off-diagonal cells, as well as for the visualisation of the sampled angle regions. In the visualisation we propose to present the entire sampled region of angles for each cell, but grey shaded according to percentiles such that more critical angles are lighter grey. In our implementation we do not present the p-values for the quantitative tests, but the p-values are available in the R-object to support the visualization if necessary. It is important to remember that there are $d \times d$ p-values, which means that attention should be given to a possible multiple testing issue.}

@@

\hl{Figure \ref{plot.simAngle} presents two examples of angle plots based on simulated data. For dataset $\mathbf{A}$, we find very short off-diagonal arrows except for the off-diagonal cells corresponding to the 3'rd and 4'th eigenvalues. But since the arrows in cells (3,4) and (4,3) lie well within the shaded angle regions (supported by p-values 0.9119 and 0.9103) this is in agreement with identical data structures, where the 3'rd and 4'th eigenvalues are close (in this situation corresponding to the two independent variables in \emph{model 1}). For dataset $\mathbf{B}$, on the other hand, we find very little agreement in all components, which should be expected, as the two groups in the dataset correspond to different data structures.}

\subsection*{The chroma plot}
The chroma plot is primarily an illustration of differences in loading patterns and it targets the question of how the roles of the original variables are different between the two datasets, thus leading the data structure comparison question back to its original, empirical context. The chroma plot consists of two panels, one for each dataset, made up of colored bars. These bars each represent a principal component and their coloring illustrates the relative weights of the $d$ original variables, that is, their absolute, normalized loading contributions. More specifically, when illustrating the $i$th principal component in the two data subsets, we plot vertical bars of length one that has been divided into $d$ segments of different colors. The width of the $j$th colored segment is given by
\begin{align*}
\omega_{ij}^x &= \frac{\lvert\eta_{ij}^x \rvert}{\sum_{k=1}^d \lvert\eta_{ik}^x\rvert}, &
\omega_{ij}^y &= \frac{\lvert\eta_{ij}^y\rvert}{\sum_{k=1}^d \lvert\eta_{ik}^y\rvert},
\end{align*}
where $\eta_{ij}^x$ and $\eta_{ij}^y$ denotes the $j$th entry of $\eta_{i}^x$ and $\eta_{i}^y$, respectively. Due to the indeterminacy of the loading signs, all the signs are removed from the coefficients in the loadings. The bars are ordered according to the eigenvalues and they are annotated with the cumulative percentage explained variance of that component, that is, the scaled and summed variance contributions,
\begin{align*}
\tilde\Sigma^x_{i} &= \frac{\sum_{j=1}^i \lambda_{j}^x}{\sum_{k=1}^d \lambda_{k}^x} = \frac{\sum_{j=1}^i \lambda_{j}^x}{d}, &
\tilde\Sigma^y_{i} &= \frac{\sum_{j=1}^i \lambda_{j}^y}{\sum_{k=1}^d \lambda_{k}^y} = \frac{\sum_{j=1}^i \lambda_{j}^y}{d}.
\end{align*}
Especially when $d$ is large, we recommend plotting only a select set of interesting principal components (for instance identified by use of the angle plot). In this scenario, the annotations should rather be the non-cumulative variance contributions, $\tilde\sigma_{i}^x = \frac{\lambda_{i}^x}{d}$ and $\tilde\sigma_{i}^y = \frac{\lambda_{i}^y}{d}$.

@@

In Figure  \ref{plot.simChroma}, two examples of chroma plots are available. For dataset $\mathbf{A}$, we see very similar visual patterns across the two panels, corresponding well with the fact that both groups are drawn from the same underlying model. For dataset $\mathbf{B}$, we see different color compositions for the bars in the left- and the right panel, illustrating that the data structures are not identical for the two groups. The plots resulting from this procedure should be inspected focusing on two properties: Similarities in loading patterns, which will correspond to similar visual impressions, and similarities in variance contributions. For each component, the loadings describe how influential the different variables are on that component. Therefore, the chroma plot allows us to make qualitative statements about the original datasets, such as \textit{"variable 1 is generally more influential in subset A than it is in subset B"}, thereby helping us to understand where and why potential data structure differences are found.

@@

Finally, we turn to a few comments about how the three plots presented thus far can be used together. We suggest that the plots are applied in the  order in which they were presented, thereby moving from overall assesments of similarity to explanations about the nature of potential data structure differences. This workflow is summarized in Figure \ref{fig:flowchart} and should follow three steps:
\begin{enumerate}
\item Make a CE plot. If this plots suggests similarity in eigenvalues then it is appropriate to move on to the angle plot to assess similarity in loadings. Otherwise, it can be concluded that the structures of the two datasets are not similar. 
\item Make an angle plot. If this plots suggests similarities in both eigenvalues and loadings, then the data structure comparison exploration is concluded. If the angle plots suggests differences in data structures, identify which components are responsible for the differences. 
\item Make a chroma plot of the components selected to be different by the angle plot. Use this information to discuss \textit{how} and \textit{why} the data structures might be different. 
\end{enumerate}
We will now turn to a data example, showing how the PCADSC tools can be used in practice.

@@

Moving on to the angle plot, we find that the differences are primarily to be found in the second, third, and fourth principal components (PCs). The blue arrows visualize the decomposition of the principal components for the Bulgarian dataset in the coordinate system of the Danish dataset. We see that PC2 also loads on PC3, that PC3 also loads on PC4, and that PC4 also loads on PC2 and PC3. The red arrows visualize the decomposition of the principal components for the Danish dataset in the coordinate system of the Bulgarian dataset. Here, we see that PC2 also loads on PC4, that PC3 also loads on PC2 and PC4, and that PC4 also loads on PC3. \hl{These statements are supported by p-values for the hypotheses of no loading that all are less than $0.002$.} Thus, if we wish to understand why differences in the data structures occur, an inspection of the loadings of components 2, 3 and 4 might be informative.

>>>>

We assume that the variables have been standardized before PCA is conducted, which is equivalent to working with the empirical correlation matrices rather than the empirical covariance matrices. We also assume that all considered variables have a numerical interpretation, for example by being continuous or ordinally categorical. Moreover, we use the following notation: 
$\mathbf{X} \in \RR^{n_x \times d} $ and $\mathbf{Y} \in \RR^{n_y \times d} $ are datasets containing the same number of variables, $d$, but possibly different numbers of observations, $n_x$ and $n_y$. We use $X_j$ to refer to the $j$th variables of $\mathbf{X}$ and $x_{ij}$ to refer to the $i$th observation  within that variable, while $x_{i \cdot}$ is the full $i$th observation row. Note that $X_j \in \RR^{n_x}$, $x_{ij} \in \RR$, and $x_{i \cdot} \in \RR^d$. We let $\bar{X} = (\bar{X}_1, ... , \bar{X}_d)^T = \frac{1}{n_x} \sum_{i = 1} x_{i \cdot}$ denote the variable averages and use $S_x =  \frac{1}{n_x - 1} \sum_{i = i}^n (x_{i \cdot} - \bar{X})(x_{i \cdot} - \bar{X})^T \in \RR^{d \times d}$ to denote the empirical covariance matrix of $\mathbf{X}$.

@@

\subsection*{The angle plot}
The angle plot compares both loadings and eigenvalues at once and it can be used to understand the information loss if the data structure of one dataset is superimposed on the other, thereby revealing which principal components (i.e. loading and eigenvalue pairs) are most similar and most different across the two datasets. Let $\lambda_{\max} = \max\{ \lambda_{1}^x, \lambda_{1}^y \}$ be the largest eigenvalue for the two datasets. Then the empirical correlation matrix for the $\mathbf{X}$ dataset, $S_x$, has the following orthogonal decomposition in the coordinate system of the $\mathbf{Y}$ dataset
\begin{equation*}
S_x = \sum_{j=1}^d \lambda_{j}^x \eta_{j}^x ({\eta_{j}^x})^\top
= \lambda_{\max} \sum_{k=1}^d
\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{k}^x}{\lambda_{\max}}} \eta_{j}^y ({\eta_{j}^y}^\top \eta_{k}^x) \Bigg)
{\Bigg( \sum_{j=1}^d \sqrt{\frac{\lambda_{k}^x}{\lambda_{\max}}} \eta_{j}^y ({\eta_{j}^y}^\top \eta_{k}^x) \Bigg)}^\top,
\end{equation*}
and we have a similar decomposition of $S_y$ in the coordinate system of the $\mathbf{X}$ dataset. We propose to visualize these two decompositions in a $d \times d$ grid display. In the $j$th row and $k$th column of this display we plot two arrows based at the lower left corner of the grid cell. The first arrow has length $\mu_{jk}$ and angle $\theta_{jk}/2$ counterclockwise from the diagonal, and the second arrow has length $\nu_{jk}$ and angle $\theta_{jk}/2$ clockwise from the diagonal. To facilitate the following description we will refer to the arrows drawn counterclockwise as the blue arrows, and the arrows drawn clockwise as the red arrows. The lengths $\mu_{jk}$ and $\nu_{jk}$, and the angle $\theta_{jk}$, are given by
\begin{align*}
\mu_{jk} &= \sqrt{\frac{\lambda_{k}^x}{\lambda_{\max}}} \lvert {\eta_{k}^x}^\top \eta_{j}^y \rvert, &
\nu_{jk} &= \sqrt{\frac{\lambda_{j}^y}{\lambda_{\max}}} \lvert {\eta_{j}^y}^\top \eta_{k}^x \rvert, &
\theta_{jk} &= \acos(\lvert {\eta_{k}^x}^\top \eta_{j}^y \rvert).
\end{align*}
For two $d$-dimensional, unit length vectors $a$ and $b$, it holds that $ a^\top b = \langle a, b \rangle = \tilde{c}(a,b)$, where $\tilde{c}$ denotes the empirical correlation function. Thus, in the angle plot, we are essentially looking at the absolute values of correlations between loadings that have been scaled according to their variance contributions. The absolute value of the projection ${\eta_{k}^x}^\top \eta_{j}^y$ is inserted due to the indeterminacy of the direction of loading vectors. This indeterminacy implies that the angle between loadings from the two datasets can always be chosen to be in the interval $[0,\pi/2]$, and hence the decomposition of $S_x$ and $S_y$ can be visualized in a joint plot by dividing the angles by two and using counterclockwise and clockwise shifts from the diagonal. Furthermore, the scaling of the lengths by $\lambda_{\max}$ is made so that the longest arrow has at most unit length.

In the \emph{angle plot}, the blue arrows in the $k$th column of the grid display visualize the decomposition of the $k$th principal component of the $\mathbf{A}$ dataset in the coordinate system of the $\mathbf{B}$ dataset. Similarly, the red arrows in the $j$th row visualize the decomposition of the $j$th principal components of the $\mathbf{B}$ dataset in the coordinate system of the $\mathbf{A}$ dataset. If we have coinciding blue and red arrows along the diagonal in the grid display, and arrows of vanishing lengths in the off-diagonal cells, then the structures of the two datasets are identical. In contrast, differences in the eigenvalues are visualized as differences in the lengths of the blue and the red arrows, also in the diagonal. And loadings in other directions than the corresponding loading from the other dataset are visualized as angle separation of the blue and the red arrows in the diagonal cells, as well as arrows of non vanishing length in the off-diagonal cells.

However, there is a generic situation where the angles between the arrows lack discriminative power, namely when two or more neighboring eigenvalues are identical or close. We refer to this situation as \textit{eigenvalue indeterminancy}. For instance, multiple eigenvalues close to 1 arise if more than one of the variables are uncorrelated, or close to uncorrelated, of all the other variables. In this situation only the corresponding eigenspace is well estimated, but the rotation of the associated eigenvectors will be subject to sampling variability. To investigate if we are in such a situation of identical data structures with neighbouring eigenvalues being close, where comparably long arrows may occur in the corresponding blocks of off-diagonal cells, we have implemented a permutation test similar as done for the CE-plot. For each random reallocation we recompute the angles $\theta_{jk}$. The resampled angles may both be used for quantitative tests, where large angles are critical for the diagonal cells and small angles are critical for the off-diagonal cells, as well as for the visualisation of the sampled angle regions. In the visualization we propose to present the entire sampled region of angles for each cell, but gray shaded according to percentiles such that more critical angles are lighter gray. %In our implementation, we do not present the $p$-values for the quantitative tests, but the $p$-values are available in the R-object to support the visualization if necessary. It is important to remember that there are $d \times d$ $p$-values, which means that attention should be given to a possible multiple testing issue.

@@

Figure \ref{plot.simAngle} presents two examples of angle plots based on simulated data. For dataset $\mathbf{A}$, we find very short off-diagonal arrows except for the off-diagonal cells corresponding to the 3rd and 4th eigenvalues. But since the arrows in cells (3,4) and (4,3) lie well within the shaded angle regions (supported by p-values 0.9119 and 0.9103) this is in agreement with identical data structures, where the 3rd and 4th eigenvalues suffer from indeterminancy (in this situation due to the two independent variables in \emph{model 1}). For dataset $\mathbf{B}$, on the other hand, we find very little agreement in all components, which should be expected, as the two groups in the dataset correspond to different data structures.

\subsection*{The chroma plot}
The chroma plot is primarily an illustration of differences in loading patterns and it targets the question of how the roles of the original variables are different between the two datasets, thus leading the data structure comparison question back to its original, empirical context. The chroma plot consists of two panels, one for each dataset, made up of colored bars. These bars each represent a principal component and their coloring illustrates the relative weights of the $d$ original variables, that is, their absolute, normalized loading contributions. More specifically, when illustrating the $i$th principal component in the two data subsets, we plot vertical bars of length one that has been divided into $d$ segments of different colors. The width of the $j$th colored segment is given by
\begin{align*}
\omega_{ij}^x &= \frac{\lvert\eta_{ij}^x \rvert}{\sum_{k=1}^d \lvert\eta_{ik}^x\rvert}, &
\omega_{ij}^y &= \frac{\lvert\eta_{ij}^y\rvert}{\sum_{k=1}^d \lvert\eta_{ik}^y\rvert},
\end{align*}
where $\eta_{ij}^x$ and $\eta_{ij}^y$ denotes the $j$th entry of $\eta_{i}^x$ and $\eta_{i}^y$, respectively. Due to the indeterminacy of the loading signs, all the signs are removed from the coefficients in the loadings. The bars are ordered according to the eigenvalues and they are annotated with the cumulative percentage explained variance of that component, that is, the scaled and summed variance contributions,
\begin{align*}
\tilde\Sigma^x_{i} &= \frac{\sum_{j=1}^i \lambda_{j}^x}{\sum_{k=1}^d \lambda_{k}^x} = \frac{\sum_{j=1}^i \lambda_{j}^x}{d}, &
\tilde\Sigma^y_{i} &= \frac{\sum_{j=1}^i \lambda_{j}^y}{\sum_{k=1}^d \lambda_{k}^y} = \frac{\sum_{j=1}^i \lambda_{j}^y}{d}.
\end{align*}
Especially when $d$ is large, we recommend plotting only a select set of interesting principal components (for instance identified by use of the angle plot). In this scenario, the annotations should rather be the non-cumulative variance contributions, $\tilde\sigma_{i}^x = \frac{\lambda_{i}^x}{d}$ and $\tilde\sigma_{i}^y = \frac{\lambda_{i}^y}{d}$. 
If there is eigenvalue indeterminancy for neighboring components, as discussed above, the loading patterns for these components will not have a meaningful interpretation. Therefore, one should not compare the chroma bars for component that were identified to suffer from eigenvalue indeterminancy by the angle plot.

@@

In Figure  \ref{plot.simChroma}, two examples of chroma plots are available. For dataset $\mathbf{A}$, we see very similar visual patterns across the two panels for components 1, 2, 5 and 6, corresponding well with the fact that both groups are drawn from the same underlying model. Components 3 and 4 do not have similar loading patterns across the two subsets of the data, but this is expected since we identified these components to suffer from eigenvalue indeterminancy in the angle plot. For dataset $\mathbf{B}$, we see different color compositions for the bars in the left- and the right panel, illustrating that the data structures are not identical for the two groups. The plots resulting from this procedure should be inspected focusing on two properties: Similarities in loading patterns, which will correspond to similar visual impressions, and similarities in variance contributions. For each component, the loadings describe how influential the different variables are on that component. Therefore, the chroma plot allows us to make qualitative statements about the original datasets, such as \textit{"variable 1 is generally more influential in subset A than it is in subset B"}, thereby helping us to understand where and why potential data structure differences are found.

@@

Finally, we turn to a few comments about how the three plots presented thus far can be used together. We suggest that the plots are applied in the  order in which they were presented, thereby moving from overall assesments of similarity to explanations about the nature of potential data structure differences. This workflow is summarized in Figure \ref{fig:flowchart} and should follow three steps:
\begin{enumerate}
\item Make a CE plot. If this plots suggests similarity in eigenvalues then it is appropriate to move on to the angle plot to assess similarity in loadings. Otherwise, it can be concluded that the structures of the two datasets are not similar. 
\item Make an angle plot. If this plots suggests similarities in both eigenvalues and loadings, then the data structure comparison exploration is concluded. If the angle plots suggests differences in data structures, identify which components are responsible for the differences. 
\item Make a chroma plot of the components selected to be different by the angle plot, but whose differences are not due to eigenvalue indeterminancy among neighboring components. Use this information to discuss \textit{how} and \textit{why} the data structures might be different. 
\end{enumerate}
We will now turn to a data example, showing how the PCADSC tools can be used in practice.

@@

Moving on to the angle plot, we find that the differences are primarily to be found in the second, third, and fourth principal components (PCs). The blue arrows visualize the decomposition of the principal components for the Bulgarian dataset in the coordinate system of the Danish dataset. We see that PC2 also loads on PC3, that PC3 also loads on PC4, and that PC4 also loads on PC2 and PC3. The red arrows visualize the decomposition of the principal components for the Danish dataset in the coordinate system of the Bulgarian dataset. Here, we see that PC2 also loads on PC4, that PC3 also loads on PC2 and PC4, and that PC4 also loads on PC3. 
%\hl{These statements are supported by p-values for the hypotheses of no loading that all are less than $0.002$.} 
Thus, if we wish to understand why differences in the data structures occur, an inspection of the loadings of components 2, 3 and 4 might be informative.
